---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The benchmark

The function uses synthetic data that is a modified version of the one found in the [Running brms models with within-chain parallelization vignette](https://cran.r-project.org/web/packages/brms/vignettes/brms_threading.html) (Weber & Bürkner, 2025). 

The model in the benchmark function is a simple Poisson MLM model with two predictors and clustering (random intercept and fixed slopes) also from the vignette. It uses 4000 iterations with the first 2000 being warmup. The model assumes that the machine has at least 4 cores and will spawn 4 chains, one for each core.

The threads argument of the function sets the number of within-chain threads per chain. At threads = 1, each MCMC chain runs on a single core (and 4 cores in total). At threads = 2, each chain's likelihood is split across 2 threads, so in total 4 x 2 = 8 cores are used. At threads = 3 it 12 cores, and so on. As a note, keep 4 x threads at or below the number of available (fast, if performance cores) physical cores.


```{r, echo = TRUE, eval = FALSE}
model_poisson <- brm(
    y ~ 1 + x1 + x2 + (1 | g),
    data = benchmark,
    family = poisson(),
    iter = 4000,
    warmup = 2000,
    chains = 4,
    cores = 4,
    threads = if (threads > 1) threading(threads) else NULL,
    seed = 1234,
    prior = prior(normal(0, 1), class = b) +
      prior(constant(1), class = sd, group = g),
    backend = "cmdstanr",
    save_pars = save_pars(all = TRUE)
  )
```

## Results

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
res <- read_csv("results.csv")

res |>
  group_by(machine, threads) |>
  summarise(mean_chain = min(mean_chain), .groups = "drop") |>
  group_by(machine) |>
  mutate(speedup = mean_chain[threads == 1] / mean_chain) |>
  ungroup() |>
  mutate(cell = sprintf("%.1f s (%.2f×)", mean_chain, speedup),
         cell = if_else(threads == 1, sprintf("%.1f s", mean_chain), cell)) |>
  select(machine, threads, cell) |>
  pivot_wider(names_from = threads, values_from = cell,
              names_prefix = "", names_glue = "{threads} thread{ifelse(threads > 1, 's', '')}") |>
  arrange(desc(`1 thread`)) |>
  knitr::kable(align = "lrrrr", caption = "Mean chain time (speed-up vs 1 thread)")
```

## The graph

```{r, echo=FALSE, message=FALSE, fig.path="figures/", fig.width=7, fig.height=4, dpi=200}

res |>
  group_by(machine, threads) |>
  summarise(mean_chain = min(mean_chain), .groups = "drop") |>
  ggplot(aes(factor(threads), mean_chain, colour = machine, group = machine)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_text(aes(label = mean_chain), vjust = -0.9, size = 3, show.legend = FALSE) +
  scale_y_continuous(limits = c(20, 60)) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "threads per chain (4 chains)",
       y = "mean chain time (in sec.)",
       colour = NULL,
       title = "brms Poisson GLMM, 10k rows") +
  see::theme_modern(base_size = 12) +
  theme(legend.position = "right")
```

## References

Weber, S., & Bürkner, P.-C. (2025). Running brms models with within-chain parallelization [Package vignette]. brms (Version 2.23.0). <https://cran.r-project.org/web/packages/brms/vignettes/brms_threading.html>
